CS182: NBA Scheduler

### 2015-12-10 paper.tex
Revised paper

    \subsection{Additional Assumptions}

    There are no non-NBA events in the venues that may further constrain a team's schedule.

@@

	We created simple classes representing teams, divisions, and conferences mimicking the structure of the league. We also created a data generation (datagen) class that generates a master list of dates - these dates conform to the scheduling requirements mentioned above. In particular, Christmas Eve is obviously December 24, we randomly select a week for an All-Star Break in February and the NCAA Men's Division I Basketball Championship is assumed to be the first Monday of April.

	Once we have our list of dates, we map them to indices ranging from 0 to the number of dates. All our subsequent calculations are based on these numbered indices. For each team, we randomly choose 3-5 dates in every 5-day period as potential home dates. We will discuss the implications of this in our results.

	\section{Integer Linear Programming}
	Integer linear programming is a mathematical optimization where all variables are restricted to integers. The objective function and the constraints are linear. In the NBA scheduling problem, the objective function is not defined and the solver has to solve the constraints as formulated in the problem. We solved this problem in 2 steps. The first was to attempt getting a schedule roundwise without considering the dates. Next we included dates and were able to create a player friendly schedule where there no more than 3 games in 5 days.

	\subsection{Methodology for the roundwise schedule}
	We create a variable for every team against every other team, for every round. In this part of the problem, we assume that there are 15 games every round with every team playing in that particular round. The first part of the variable is the home team followed by the away team and the round number.

	\begin{eqnarray*}
		variable : (Home \ Team, Away \ Team, Round \ Number)
	\end{eqnarray*}

	We introduce the following constraints to the ILP.
	\begin{enumerate}
		\item All teams play 82 games, 41 home and away
		\item All constraints for the division, in conference and out of conference teams
		\item No team can play more than one game in a round
		\item Team can only play either home or away
	\end{enumerate}
	The above ILP was solved using the PuLP as well as the GUROBI solver. The PuLP solver took 40 minutes to converge to solution whereas the GUROBI solver took 6 seconds on average. The schedule was verified using tests to check if all constraints were satisfied.

	\subsection{Generating the player friendly schedule}
	Using the dates generated by datagen, we formulated the ILP for a player friendly schedule. The variables are created for every date instead of every round.  To reduce the number of variables we generate the variables for every possible home and away dates only. For every team, we create the variables for every home date for every opposition. Similarly, we create the variables for every away date for a particular opponent.

	\begin{eqnarray*}
		variable : (Home \ Team, Away \ Team, Game \ Date)
	\end{eqnarray*}

	This reduces the number of variables by about 70\%. The time taken to create the constraints and to solve the ILP are greatly reduced. We use the following constraints.
	\begin{enumerate}
		\item All teams play 82 games, 41 home and away
		\item All constraints for the division, in conference and out of conference teams
		\item No team can play more than one game on a single date
		\item Team can only play either home or away
		\item A team can only play 3 or lesser matches in 5 days
		\item A team does not play 2 away games without a day's break
		\item A team does not play an away game right after a home game
	\end{enumerate}
	The above ILP was solved using the PuLP as well as the GUROBI solver. The PuLP solver took 1 hour to converge to solution whereas the GUROBI solver took 29 seconds on an average. In some cases the 4 games in 5 days was unavoidable due to mismatch in dates, but our solution aimed to minimize it.

@@

	\subsection{Matchups CSP}
	Within our code, for each team we initialize a list of opposing teams with each team repeated to the number of times these two teams play. Our $MSP$ then will form the unique list of teams from this list to try assignments.

@@

	\subsubsection{Heuristics}
	There were a number of heuristics we used to solve this CSP.
	\begin{enumerate}
		\item We applied forward checking to each successor state we looked at. This consisted of checking that the min\_count for every team was at least 1 and the sum\_count over every pair of teams was at least as large as the number of games they had left to play.
		\item We chose the next team/game to assign by finding the team with the lowest min\_count and it's associated game.
		\item After choosing $t,g$ for the team/game number to next assign, we ordered the possible opponents using the least-constrained heuristic. This involved trying out the assignment and recalculating min\_count($t$). We then ordered the possible opponents in decreasing order - the opponent that kept the min\_count from dropping the most was the least constrained assignment.
	\end{enumerate}

@@

	\subsection{Dates/Venues CSP}
	Using the results from the Matchups CSP, we formulated this problem first as a DFS again. In our code, for any pair of teams we stored a list of True/False tuples according to the number of times those two teams played. If the teams played three times, there'd be two Trues and two Falses. We also created a master list of possible dates with the keys being $(Team_1,Team_2,game\_num,True/False)$. This represented the possible dates for any matchup and home/away combination (True meaning $Team_1$ was home). For the home team, let the set of home dates generated in our initial data generation process be $D$. We defined a hyperparameter $window$, and assigned the values to this dictionary as:

	\begin{eqnarray*}
		\forall x \in D: x \geq 2*game\_num - window \land x \leq 2*game\_num + window
	\end{eqnarray*}

	The rationale here is the season is typically 160-170 days, meaning ideally each game is every other day. So, we specify the domain of possible dates for a game to be a window of dates centered at that ideal date. The heuristics used were as follows:
	\begin{enumerate}
		\item Forward checking was relatively straightforward in this CSP. We remove any True/False possibilities for any pair of teams if one team has reached the limit of home or away games for that matchup. For any date $d$ we assign for some game\_num, we remove the dates prior to $d$ for any game after game\_num, and any date after $d$ for any game before game\_num for any matchup involving one of these teams.
		\item Choosing what matchup to next assign was the most perplexing part of this problem. We initially chose the team with the most assignments up to that point - essentially trying to assign all home/away matchups for a specific team at a time. The difficulty in this CSP is finding the right assignments for matchups of teams that play 3 times. There's an added complexity of having the right combination of 1 home/2 away and 2 home/1 away matchups that keep each team's total home/away games even. What was originally happening was the DFS would assign every matchup for every team until the last few, but it was impossible to find a right assignment for them because too many teams had their 3-game matchups with a given 2-home or 2-aways. There wasn't a good backtracking method implemented once that happened, so the CSP would stagnate through all possible iterations. Given that, we decided to first assign all matchups between opponents that play an even number of times and only after all those were completed do we start on the 3-game matchups. This was the key to solving this CSP.
		\item Once we had our matchup we wanted to assign, we checked which values True or False were possible - if a team had reached their limit of home/away games then there was no ordering to be done. If both teams could potentially be the home team, we chose to first try the team with the fewer number of home games be the potential home team. Then, for any Home/Away assignment, we had our master dates list with potential dates. We ordered these according to the distance to the ``ideal date'', the 2*game\_num we had used to first assign the window of possible dates. The resulting list of possible assignments was interleaved True/False assignments with their associated dates based on distance to the ``ideal date''.
		\begin{itemize}
			\item As an example suppose for $(Team_1,Team_2,game\_num=10)$, the window of home dates for $Team_1$ are $(17,19,21,23)$ and the window of home dates for $Team_2$ are $(18,20)$. We want to first try $Team_2$ as the home team because to that point it's had less home assignments, so False would take priority to True. The resulting ordering of the domain would be: (20,False), (19,True), (18,False), (21,True), (23,True), (17,True) - ordering of 19/21 and 17/23 could be flipped based Python's sort implementation.
		\end{itemize}
	\end{enumerate}

	There's potentially no solution to our problem given the availability of home dates. This goes back to our original dates formulation - randomly selecting between 3-5 home dates every 5 days of the season for every team. If every date were possible - i.e. each team's possible home dates were every day of the season, our CSP solves in under a minute with typically under 1000 states explored. If we set the possible home dates to be 3 out of every 5 dates, our CSP solved roughly half the time. This was the rationale for us to randomly choose between 3-5 home dates, as we generally obtain solutions and they're more interesting than a game every other day.

>>>>

    \subsection{Additional Assumptions}
    There are no non-NBA events in the venues that may further constrain a team's schedule.

@@

	We created simple classes representing teams, divisions, and conferences mimicking the structure of the league. We also created a data generation (datagen) class that generates a master list of dates that conform to the scheduling requirements mentioned above. In particular, Christmas Eve is obviously December 24, we randomly select a week for an All-Star Break in February and the NCAA Men's Division I Basketball Championship is assumed to be the first Monday of April.

	Once we have our list of dates, we map them to indices ranging from 0 to the total number of dates. All our subsequent calculations are based on these numbered indices. For each team, we randomly choose 3-5 dates in every 5-day period as potential home dates. We will discuss the implications of this in our results.

	\section{Integer Linear Programming}
	Integer linear programming is a mathematical optimization where all variables are restricted to integers. The objective function and the constraints are linear equations. In the NBA scheduling problem, the objective function is not defined and the solver has to solve the constraints as formulated in the problem. We solved this problem in 2 steps. The first step was picking the home team, away team, and game number for each matchup, without considering dates.  For this step, if Team 1 and Team 2 play each other in game number $x$, it is the Team 1's game $x$ and Team 2's game $x$, as opposed to the game being, for instance, game 3 for Team 1 and game 4 for Team 2.  Once we verified correct results for generating matchups, the second step was to modify the problem to include dates.  After completing the second problem, we created a player friendly schedule where there no more than 3 games in 5 days.

	\subsection{Methodology for the assigning game numbers to matchups}
	We create a variable all combinations of Team 1, Team 2, game number. It is important to note that every team plays one game for each game number (1 through 82).

	\begin{eqnarray*}
		variable : (Home \ Team, Away \ Team, Game Number)
	\end{eqnarray*}

	We introduce the following constraints to the ILP.
	\begin{enumerate}
		\item All teams play 82 games, 41 home and away
		\item All constraints for the division, in conference and out of conference teams
		\item No team can be involved in more than one matchup per game number
		\item Team can only play either home or away
	\end{enumerate}
	The above ILP was solved using the PuLP as well as the GUROBI solver. The PuLP solver took 40 minutes to converge to solution whereas the GUROBI solver took 6 seconds on average. The schedule was verified using tests to check if all constraints were satisfied.

	\subsection{Generating the player friendly schedule}
	Using the possible home dates for each team generated by datagen, we formulated the ILP for a player friendly schedule.  Note that the difference here is that we created variables for every date instead of every game number.  To reduce the number of variables, we only created variables using the possible home dates provided by the home team. For each pair of teams ($Team_1, Team_2$), we include variables for the possiblity that Team 1 is home and Team 2 is away and vice versa.

	\begin{eqnarray*}
		variable : (Home \ Team, Away \ Team, Date)
	\end{eqnarray*}

	Creating variables using only the home dates provided by the home team instead of all possible dates reduces the number of variables by about 70\%. The optimization greatly reduced the runtime of the solver . We use the following constraints.
	\begin{enumerate}
		\item All teams play 82 games, 41 home and away
		\item All constraints for the division, in conference and out of conference teams
		\item No team can play more than one game on a single date
		\item Team can only play either home or away
		\item A team can only play no more than 3 matches in 5 days
		\item A team does not play 2 away games without a day's break
		\item A team does not play an away game right after a home game
	\end{enumerate}
	The above ILP was solved using the PuLP as well as the GUROBI solver. The PuLP solver took 1 hour to converge to solution whereas the GUROBI solver took 29 seconds on an average. In some cases the 4 games in 5 days was unavoidable due to mismatch in dates, but our solution aimed to minimize it.

@@

	\subsection{Matchups CSP}
	Within our code, for each team we initialize a list of opposing teams with each team repeated the number of times the two teams play. Our $MSP$ then will form the unique list of teams from this list to try assignments.

@@

	\subsubsection{Heuristics}
	There were a number of heuristics we used to solve this CSP.
	\begin{enumerate}
		\item We applied forward checking to each successor state we looked at. This consisted of checking that the min\_count for every team was at least 1 and the sum\_count over every pair of teams was at least as large as the number of games they had left to play.
		\item We chose the next team to assign by finding the team with the lowest min\_count.
		\item After choosing $t$ as the team to assign next, we ordered the possible opponents using the least-constrained heuristic. This involved trying out the assignment and recalculating min\_count($t$). We then ordered the possible opponents in decreasing order - the opponent that kept the min\_count from dropping the most was the least constrained assignment.
		\item After choosing the two teams, we selected the matchup between the two chosen teams with the lowest game number to be the next matchup to be assigned a date
	\end{enumerate}

@@

	\subsection{Dates/Venues CSP}
	Using the results from the Matchups CSP, we formulated this problem first as a DFS again. In our code, for any pair of teams we stored a list of True/False tuples according to the number of times those two teams played. If the teams played three times, there would be two Trues and two Falses. We also created a master list of possible dates with the keys being $(Team_1,Team_2,game\_num,True/False)$. This represented the possible dates for any matchup and home/away combination (True meaning $Team_1$ was home). For the home team, let the set of home dates generated in our initial data generation process be $D$. We defined a hyperparameter $window$ and assigned the values to this dictionary as follows, where $g$ is the game number:

	\begin{eqnarray*}
		\forall x \in D: x \geq 2*g- window \land x \leq 2*g + window
	\end{eqnarray*}

	The rationale here is the season is typically between 160 and 170 days, meaning ideally each game is every other day. So, we specify the domain of possible dates for a game to be a window of dates centered at that ideal date. The heuristics used were as follows:
	\begin{enumerate}
		\item Forward checking was relatively straightforward in this CSP. We remove any True/False possibilities for any pair of teams if one team has reached the limit of home or away games for that matchup. For any date $d$ we assign for some game number $g$, we remove the dates prior to $d$ for any game after $g$ and any date after $d$ for any game before $g$ for any matchup involving one of these teams.
		\item Choosing what matchup to next assign was the most perplexing part of this problem. We initially chose the team with the most assignments up to that point - essentially trying to assign all home/away matchups for a specific team at a time. The difficulty in this CSP is finding the right assignments for matchups of teams that play 3 times. There's an added complexity of having the right combination of 1 home/2 away and 2 home/1 away matchups that keep each team's total home/away games even. What was originally happening was the DFS would assign every matchup for every team until the last few, but it was impossible to find a right assignment for them because too many teams had their 3-game matchups with a given 2-home or 2-aways. There wasn't a good backtracking method implemented once that happened, so the CSP would stagnate through all possible iterations. Given that, we decided to first assign all matchups between opponents that play an even number of times and only after all those were completed do we start on the 3-game matchups. This was the key to solving this CSP.
		\item Once we selected matchup to assign, we checked which values of True or False were possible: that is, which of the two teams could be the home team.  If a team had reached their limit of home games or away games then there is no ordering to be done because there is at most one option. If both teams could potentially be the home team, we chose to first try the team with the fewer number of home games be the potential home team. Then, for any Home/Away assignment, we had our master dates list with potential dates. We ordered these according to the distance to the ``ideal date'',  2*$g$, where $g$ is the game number we had used to first assign the window of possible dates. The resulting list contained possible assignments: alternating True/False assignments with their associated dates ordered by distance to the ``ideal date''.
		\begin{itemize}
			\item As an example suppose for $(Team_1,Team_2,$g$ =10)$, the window of home dates for $Team_1$ are $(17,19,21,23)$ and the window of home dates for $Team_2$ are $(18,20)$. We want to first try $Team_2$ as the home team because to that point it's had less home assignments, so False would take priority to True. The resulting ordering of the domain would be: (20,False), (19,True), (18,False), (21,True), (23,True), (17,True) - ordering of 19/21 and 17/23 could be flipped based Python's sort implementation.
		\end{itemize}
	\end{enumerate}

	There's potentially no solution to our problem given the availability of home dates. This goes back to our original dates formulation - randomly selecting between 3 and 5 home dates every 5 days of the season for every team. If every date were possible - i.e. each team's possible home dates were every day of the season, our CSP solves in under a minute with typically under 1000 states explored. If we set the possible home dates to be 3 out of every 5 dates, our CSP solved roughly half the time. This was the rationale for us to randomly choose 3 to 5 home dates, as we generally obtain solutions and they're more interesting than a game every other day.
